### 2021-01-02 written/eLifeRevision/Bittner_eLife2020.tex
latest revision

% What's wrong with existing methodology:
Existing approaches to the inverse problem in theoretical neuroscience fall short in three key ways. 
First, theoretical models of neural computation aim to reflect a complex biological reality, and as a result, such models lack tractable likelihoods.
Thus, standard approaches from statistical inference are unavailable.
The parameter sets obtained from likelihood-free ABC lack a formalized link to Bayesian inference (except in the unrealistic $0$-distance scenario), and lack parameter probabilities.
Second is the undesirable trade-off between the flexibility and tractability of the approximated posterior distribution.
While sampling-based approaches like ABC and Markov chain Monte Carlo (MCMC) can produce flexible posterior approximations, they must be run continually for increasing samples.
While VAE approaches can result in tractable posterior sampling and sensitivity measurements post-optimization, existing approaches have relied on simplified classes of distributions, which restrict the flexibility of the posterior approximation.
And third, you can never make assumptions of what inferred model parameters will predict.
This is well understood when considering Box's loop and the role of posterior predictive checks in the development and critique of scientific models \cite{gelman2013philosophy, blei2014build}.
Uncertainty about the properties of inferred model predictions introduce a conceptual degree of freedom to the inverse problem that may be unnecessary and undesirable given the scientific motivation.

@@

Equipped with this methodology, we then investigated three models of current importance in theoretical neuroscience.
These models were chosen to demonstrate generality through ranges of biological realism (from conductance-based biophysics to recurrent neural networks), neural system function (from pattern generation to abstract cognitive function), and network scale (from four to 100 neurons).
%First, we use EPI to produce a set of verifiable hypotheses of input-responsivity in a four neuron-type dynamical model of primary visual cortex; we then validate these hypotheses in the model.
First, we use EPI to understand the characteristics of noise that govern Fano factor in a stochastic four neuron-type model of primary visual cortex.
%Second, we demonstrated how the systematic application of EPI to levels of task performance can generate experimentally testable hypotheses regarding connectivity in superior colliculus.  
Second, we discover connectivity patterns in superior colliculus resilient to optogenetic perturbation by using EPI to condition on rapid task switching.
The novel scientific insights offered by EPI contextualize and clarify the previous studies exploring these models \cite{litwin2016inhibitory, duan2018collicular}.
%Third, we use EPI to uncover the sources of error in a low-rank recurrent neural network executing a simple mathematical task.  
Third, we emphasize the methodological advancement of EPI by inferring high-dimensional distributions of RNN connectivities exhibiting stable amplification.
These results point to the value of deep inference for the interrogation of biologically relevant models.

@@

\textbf{C}. The emergent property of intermediate hub frequency, in which the hb neuron fires at a rate between the fast and slow frequencies.

@@

@@

@@

\label{fig:STG_EPI}

@@

%Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}A gray box) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution shown in Figure \ref{fig:STG}B.
Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG_EPI}D) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution  EPI (Fig. \ref{fig:STG}E).  
This distribution -- represented for clarity as samples from the distribution -- is then a scientifically meaningful and mathematically tractable object.  
In the STG model, this distribution can be specifically queried to reveal the prototypical parameter configuration for network syncing (the mode; Figure \ref{fig:STG}E yellow star), and how network syncing decays based on changes away from the mode.
The eigenvectors (of the Hessian of the distribution at the mode) quantitatively formalize the robustness of unified intermediacy (Fig. \ref{fig:STG}B solid ($v_1$) and dashed  ($v_2$) black arrows).
Indeed, samples equidistant from the mode along these EPI-identified dimensions of sensitivity ($v_1$) and degeneracy ($v_2$) agree with error contours (Fig. \ref{fig:STG}B contours) and have diminished or preserved network syncing, respectively (Fig. \ref{fig:STG}F activity traces, Fig. S TODO) (see Section \ref{methods_STG}).

\subsection{A deep generative modeling approach to emergent property inference} \label{results_dgm}
Emergent property inference (EPI) systematizes the three-step procedure of the previous section.
%First, we consider the model as a coupled set of differential (and potentially stochastic) equations \cite{gutierrez2013multiple}.  
First, we consider the model as a coupled set of stochastic differential equations \cite{gutierrez2013multiple}.  
In the running STG example, the model activity $\mathbf{x} = \left[ x_{\text{f1}}, x_{\text{f2}}, x_{\text{hub}}, x_{\text{s1}}, x_{\text{s2}} \right]$ is the membrane potential for each neuron, which evolves according to the biophysical conductance-based equation:
\begin{equation} 
\begin{split}
C_m \frac{d\mathbf{x}(t)}{dt} = -\mathbf{h}(\mathbf{x}(t); \mathbf{z}) + d\mathbf{B}
 \end{split}
\end{equation} 
where $C_m$=1nF, and $\mathbf{h}$ is a sum of the leak, calcium, potassium, hyperpolarization, electrical, and synaptic currents, all of which have their own complicated dependence on $\mathbf{x}$ and $\mathbf{z} = [g_{\text{el}}, g_{\text{synA}}]$, and $d\mathbf{B}$ is white gaussian noise (see Section \ref{methods_STG}).

%Second, we define the emergent property, which as above is network syncing: oscillation of the entire population at an intermediate frequency of our choosing (Figure \ref{fig:STG}A bottom).
Second, we define the emergent property, which as above is ``intermediate hub  frequency" (Figure \ref{fig:STG}C).
%Quantifying this phenomenon is straightforward: we define network syncing to be that each neuron's spiking frequency -- denoted $\omega_{\text{f1}}(\mathbf{x})$, $\omega_{\text{f2}}(\mathbf{x})$, etc. -- is close to an intermediate frequency of 0.53Hz.  
%Mathematically, we achieve this via constraints on the mean and variance of $\omega_\alpha(\mathbf{x})$ for each neuron $\alpha \in \{ \text{f1}, \text{f2}, \text{hub}, \text{s1}, \text{s2} \}$:
Quantifying this phenomenon is straightforward: we stipulate that the hub neuron's spiking frequency -- denoted $\omega_{\text{hub}}(\mathbf{x})$is close to an intermediate frequency of 0.55Hz.  
Mathematically, we achieve this via constraints on the mean and variance of the hub neuron spiking frequency.

@@

Third, we perform emergent property inference: we find a distribution over parameter configurations $\mathbf{z}$, and insist that samples from this distribution produce the emergent property; in other words, they obey the constraints introduced in Equation \ref{eq:EP_STG}.  
This distribution will be chosen from a family of probability distributions $\mathcal{Q} = \left\{ q_{\bm{\theta}}(\mathbf{z}) : \bm{\theta} \in \Theta \right\}$, defined by a deep generative distribution of the normalizing flow class \cite{rezende2015variational, dinh2016density, papamakarios2017masked} -- neural networks which transform a simple distribution into a suitably complicated distribution (as is needed here).  
This deep distribution is represented in Figure \ref{fig:STG}C (see Section \ref{methods_EPI}).  
Then, mathematically, we must solve the following optimization program: 
%\begin{equation} \label{eq:EPI}
%\begin{split}
%&\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
%&\text{  s.t.  } \mathbb{E}_{\mathbf{z} \sim q_{\bm{\theta}}}\left[ \mathbb{E}_{\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})}\left[T(\mathbf{x})\right] \right] = \bm{\mu}, \\
%\end{split}
%\end{equation}
\begin{equation} \label{eq:EPI}
\begin{split}
q_{\bm{\theta}}(\mathbf{z} \mid \mathcal{X}) ~~=~~ &\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
&\text{  s.t.  } ~\mathcal{X}:~~ \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2 \\
\end{split}
\end{equation}
%where $T(\mathbf{x}), \bm{\mu}$ are defined as in Equation \ref{eq:EP}, and $p(\mathbf{x} | \mathbf{z})$ is the intractable distribution of data from the model, $\mathbf{x}$, given that model's parameters $\mathbf{z}$ (we access samples from this distribution by running the model forward).
where $f(\mathbf{x}, \mathbf{z}), \bm{\mu}$, and $\bm{\sigma}$ are defined as in Equation \ref{eq:EP}.
%The purpose of each element in this program is detailed in Figure \ref{fig:STG}D.
Finally, we recognize that many distributions in $\mathcal{Q}$ will respect the emergent property constraints, so we select that which has maximum entropy.
This principle, captured in Equation \ref{eq:EPI} by the primal objective $H$, identifies parameter distributions with minimal assumptions beyond some chosen structure \cite{jaynes1957information, elsayed2017structure, loaiza2017maximum, savin2017maximum}.
Such a normative principle of maximum entropy, which is also that of Bayesian inference, naturally fits with our scientific objective of reasoning about parametric sensitivity and robustness. 
The recovered distribution of EPI is as variable as possible along each parametric manifold such that it produces the emergent property.

@@

In relation to broader methodology, inspection of the EPI objective reveals a natural relationship to posterior inference.
%Specifically, EPI executes variational inference in an exponential family model, the sufficient statistics and mean parameter of which are defined by the emergent property statistics and values, respectively (see Section \ref{methods_VI}).
%Specifically, EPI executes variational inference in an exponential family model, the sufficient statistics and mean parameter of which are defined by $f(\mathbf{x})$, $\bm{\mu}$ and $\bm{\sigma}$ (see Section \ref{methods_VI}).
Specifically, EPI (TODO insert interpretation).
%Equipped with this method, we may examine structure in such inferred distributions or make comparisons between inferred distributions at graded values of emergent property statistics.
Equipped with this method, we may examine structure in posterior distributions or make comparisons between posteriors conditioned at different levels of the same emergent property statistic.
We now prove out the value of EPI by using it to investigate and produce novel insights about three prominent models in neuroscience.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Comprehensive input-responsivity in a nonlinear sensory system} \label{results_V1}

\begin{figure}
\begin{center}
\includegraphics[scale=0.54]{figures/fig2/fig2.pdf}
\end{center}
\caption{\small Hypothesis generation through EPI in a V1 model.  
A. Four-population model of primary visual cortex with excitatory (black), parvalbumin (blue), somatostatin (red), and VIP (green) neurons.   
Some neuron-types largely do not form synaptic projections to others  (excitatory and inhibitory projections filled and unfilled, respectively).  
B. Linear response predictions become inaccurate with greater input strength.  V1 model simulations for input (solid) $h=b$ and (dashed) $h = b + dh$. 
Stars indicate the linear response prediction.  
C. EPI distributions on differential input $dh$ conditioned on differential response $\mathcal{B}(\alpha, y)$. 
Supporting evidence for the four generated hypotheses are indicated by gray boxes with labels H1, H2, H3, and H4. 
The linear prediction from two standard deviations away from $y$ (from negative to positive) is overlaid in magenta (very small, near origin). }
 \label{fig:V1_EPI}
\end{figure}

Dynamical models of excitatory (E) and inhibitory (I) populations with supralinear input-output function have succeeded in explaining a host of experimentally documented phenomena.
In a regime characterized by inhibitory stabilization of strong recurrent excitation, these models give rise to paradoxical responses \cite{tsodyks1997paradoxical}, selective amplification  \cite{murphy2009balanced}, surround suppression \cite{ozeki2009inhibitory} and normalization \cite{rubin2015stabilized}. 
Despite their strong predictive power, E-I circuit models rely on the assumption that inhibition can be studied as an indivisible unit. 
However, experimental evidence shows that inhibition is composed of distinct elements -- parvalbumin (P), somatostatin (S), VIP (V) -- composing 80\% of GABAergic interneurons in V1 \cite{markram2004interneurons, rudy2011three, tremblay2016}, and that these inhibitory cell types follow specific connectivity patterns (Fig. \ref{fig:V1_EPI}A) \cite{pfeffer2013inhibition}.
Recent theoretical advances \cite{litwin2016inhibitory, GarciaDelMolino2017, Chen2019},  have only started to address the consequences of this multiplicity in the dynamics of V1, strongly relying on linear theoretical tools. 
Here, we go beyond linear theory by systematically generating and evaluating hypotheses of circuit model function using EPI distributions of neuron-type inputs producing various neuron-type population responses.

Specifically, we consider a four-dimensional circuit model with dynamical state given by the firing rate $x$ of each neuron-type population $x = \left[x_E, x_P , x_S, x_V \right]^\top$. 
Given a time constant of $\tau = 20$ ms and a power $n = 2$, the dynamics are driven by the rectified and exponentiated sum of recurrent ($Wx$) and external $h$ inputs:

\begin{equation}
\tau \frac{dx}{dt} = -x + [W x+ h]_+^n.
\end{equation}

We considered fixed effective connectivity weights $W$ approximated from experimental recordings of publicly available datasets of mouse V1 \cite{allen2018layer, billeh2019systematic} (see Section \ref{methods_V1}).
The input $h = b + dh$ is comprised of a baseline input  $b = \left[ b_E, b_P , b_S , b_V \right]^\top$ and a differential input $dh = \left[ dh_E , dh_P , dh_S , dh_V\right]^\top$ to each neuron-type population.  
Throughout subsequent analyses, the baseline input is $b = \left[ 1 ,1,1,1\right]^\top$.

With this model, we are interested in the differential responses of each neuron-type population to changes in input $dh$. 
Initially, we studied the linearized response of the system to input $\frac{dx_{ss}}{dh}$ at the steady state response $x_{ss}$, i.e. a fixed point. 
All analyses of this model consider the steady state response, so we drop the notation $ss$ from here on.
While this linearization accurately predicts differential responses $dx = \left[ dx_{E} , dx_{P} , dx_{S} ,dx_{V} \right]^\top$  for small differential inputs to each population $dh = \left[ 0.1 , 0.1 , 0.1 , 0.1 \right]^\top$ (Fig \ref{fig:V1_EPI}B left), the linearization is a poor predictor in this nonlinear model more generally (Fig. \ref{fig:V1_EPI}B right).  
Currently available approaches to deriving the steady state response of the system are limited.

To get a more comprehensive picture of the input-responsivity of each neuron-type beyond linear theory, we used EPI to learn a distribution of the differential inputs to each population $dh$ that produce an increase of $y$ in the rate of each neuron-type population $\alpha \in \{E, P, S, V \}$.  
We want to know the differential inputs $dh$ that result in a differential steady state $dx_{\alpha}$ (the change in $x_{\alpha}$ when receiving input $h=b + dh$ with respect to the baseline $h = b$) of value $y$ with some small, arbitrarily chosen amount of variance  $0.01^2$.   
These statements amount to the emergent property 
\begin{equation}
\mathcal{B}(\alpha, y) ~~\triangleq~~ 
\mathbb{E} \begin{bmatrix} dx_{\alpha} \\ (dx_{\alpha} - y)^2 \end{bmatrix} ~~=~~ \begin{bmatrix} y \\ 0.01^2 \end{bmatrix}.
\end{equation}
We maintain the notation $\mathcal{B}(\cdot)$ throughout the rest of the study as short hand for emergent property, which represents a different signature of computation in each application.

Using EPI, we inferred the distribution of $dh$ shown in Figure \ref{fig:V1_EPI}C producing $\mathcal{B}(\alpha, y)$.  
Columns correspond to inferred distributions of excitatory ($\alpha=E$, red), parvalbumin ($\alpha=P$, blue), somatostatin ($\alpha=S$, red) and VIP ($\alpha=V$, green) neuron-type response increases, while each row corresponds to increase amounts of $y \in \{0.1, 0.5\}$.  For each pair of parameters, we show the two-dimensional marginal distribution of samples colored by $\log q_{\bm{\theta}}(dh \mid \mathcal{B}(\alpha, y))$.  The inferred distributions immediately suggest four hypotheses: \\

{\addtolength{\leftskip}{10 mm}
H1: as is intuitive, each neuron-type's firing rate should be sensitive to that neuron-type's direct input (e.g. Fig. \ref{fig:V1_EPI}C H1 gray boxes indicate low variance in $dh_E$ when $\alpha=E$. Same observation in all inferred distributions); \\
H2: the E- and P-populations should be largely unaffected by input to the V-population (Fig. \ref{fig:V1_EPI}C H2 gray boxes indicate high variance in $dh_V$ when $\alpha \in \{E, P \}$); \\
H3: the S-population should be largely unaffected by input to the P-population  (Fig. \ref{fig:V1_EPI}C H3 gray boxes indicate high variance in $dh_P$ when $\alpha =S$); \\
H4: there should be a nonmonotonic response of the V-population with input to the E-population (Fig. \ref{fig:V1_EPI}C H4 gray boxes indicate that negative $dh_E$ should result in small $dx_{V}$, but positive $dh_E$ should elicit a larger $dx_{V}$);

>>>>

% What's wrong with existing methodology:
Existing approaches to the inverse problem in theoretical neuroscience fall short in three key ways. 
First, theoretical models of neural computation aim to reflect a complex biological reality, and as a result, such models lack tractable likelihoods.
Thus, standard approaches from statistical inference are unavailable.
The parameter sets obtained from likelihood-free ABC lack a formalized link to Bayesian inference (except in the unrealistic $0$-distance scenario), and lack parameter probabilities.
Second is the undesirable trade-off between the flexibility and tractability of the approximated posterior distribution.
While sampling-based approaches like ABC and Markov chain Monte Carlo (MCMC) can produce flexible posterior approximations, they must be run continually for increasing samples.
While VAE approaches can result in tractable posterior sampling and sensitivity measurements post-optimization, existing approaches have relied on simplified classes of distributions, which restrict the flexibility of the posterior approximation.
And third, one can never assume what inferred model parameters may predict.
This is well understood when considering Box's loop and the role of posterior predictive checks in the development and critique of scientific models \cite{gelman2013philosophy, blei2014build}.
Uncertainty about the properties of inferred model predictions introduce a conceptual degree of freedom to the inverse problem that may be unnecessary and undesirable given the scientific motivation.

@@

Equipped with this methodology, we then investigated three models of current importance in theoretical neuroscience.
These models were chosen to demonstrate generality through ranges of biological realism (from conductance-based biophysics to recurrent neural networks), neural system function (from pattern generation to abstract cognitive function), and network scale (from four to hundreds of neurons).
%First, we use EPI to produce a set of verifiable hypotheses of input-responsivity in a four neuron-type dynamical model of primary visual cortex; we then validate these hypotheses in the model.
First, we use EPI to understand the characteristics of noise that govern Fano factor in a stochastic four neuron-type model of primary visual cortex.
%Second, we demonstrated how the systematic application of EPI to levels of task performance can generate experimentally testable hypotheses regarding connectivity in superior colliculus.  
Second, we discover connectivity patterns in superior colliculus resilient to optogenetic perturbation by using EPI to condition on rapid task switching.
The novel scientific insights offered by EPI contextualize and clarify the previous studies exploring these models \cite{litwin2016inhibitory, duan2018collicular}.
%Third, we use EPI to uncover the sources of error in a low-rank recurrent neural network executing a simple mathematical task.  
Third, we emphasize the methodological advancement of EPI by inferring high-dimensional distributions of RNN connectivities exhibiting stable amplification.
These results point to the value of deep inference for the interrogation of biologically relevant models.

@@

\textbf{C}. The emergent property of intermediate hub frequency, in which the hub neuron fires at a rate between the fast and slow frequencies.

@@

In EPI optimization, stochastic gradient steps in $\bm{\theta}$ are taken such that entropy is maximized, and the emergent property $\mathcal{X}$ is produced.

@@

The black and gray dashed lines show the mean and two standard deviations according the emergent property, respectively.

@@

 \label{fig:STG}

@@

%Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}A gray box) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution shown in Figure \ref{fig:STG}B.
Before presenting technical details (in the following section), let us understand emergent property inference schematically:  EPI (Fig. \ref{fig:STG}D) takes, as input, the model and the specified emergent property, and as its output, produces the parameter distribution  EPI (Fig. \ref{fig:STG}E).  
This distribution -- represented for clarity as samples from the distribution -- is then a scientifically meaningful and mathematically tractable object.  
In the STG model, this distribution can be specifically queried to reveal the prototypical parameter configuration for network syncing (the mode; Figure \ref{fig:STG}E yellow star), and how network syncing decays based on changes away from the mode.
The eigenvectors (of the Hessian of the distribution at the mode) quantitatively formalize the robustness of unified intermediacy (Fig. \ref{fig:STG}B solid ($v_1$) and dashed  ($v_2$) black arrows).
Indeed, samples equidistant from the mode along these EPI-identified dimensions of sensitivity ($v_1$) and degeneracy ($v_2$) agree with error contours (Fig. \ref{fig:STG}B contours) and have diminished or preserved network syncing, respectively (Fig. \ref{fig:STG}F activity traces, Fig. S TODO) (see Section \ref{methods_STG}).

\subsection{A deep generative modeling approach to emergent property inference} \label{results_dgm}
Emergent property inference (EPI) systematizes the three-step procedure of the previous section.
%First, we consider the model as a coupled set of differential (and potentially stochastic) equations \cite{gutierrez2013multiple}.  
First, we consider the model as a coupled set of differential equations \cite{gutierrez2013multiple}.  
In the running STG example, the model activity $\mathbf{x} = \left[ x_{\text{f1}}, x_{\text{f2}}, x_{\text{hub}}, x_{\text{s1}}, x_{\text{s2}} \right]$ is the membrane potential for each neuron, which evolves according to the biophysical conductance-based equation:
\begin{equation} 
\begin{split}
C_m \frac{d\mathbf{x}(t)}{dt} = -h(\mathbf{x}(t); \mathbf{z}) + d\mathbf{B}
 \end{split}
\end{equation} 
where $C_m$=1nF, and $\mathbf{h}$ is a sum of the leak, calcium, potassium, hyperpolarization, electrical, and synaptic currents, all of which have their own complicated dependence on $\mathbf{x}$ and $\mathbf{z} = [g_{\text{el}}, g_{\text{synA}}]$, and $d\mathbf{B}$ is white gaussian noise (see Section \ref{methods_STG}).

%Second, we define the emergent property, which as above is network syncing: oscillation of the entire population at an intermediate frequency of our choosing (Figure \ref{fig:STG}A bottom).
Second, we define the emergent property, which as above is ``intermediate hub  frequency" (Figure \ref{fig:STG}C).
%Quantifying this phenomenon is straightforward: we define network syncing to be that each neuron's spiking frequency -- denoted $\omega_{\text{f1}}(\mathbf{x})$, $\omega_{\text{f2}}(\mathbf{x})$, etc. -- is close to an intermediate frequency of 0.53Hz.  
%Mathematically, we achieve this via constraints on the mean and variance of $\omega_\alpha(\mathbf{x})$ for each neuron $\alpha \in \{ \text{f1}, \text{f2}, \text{hub}, \text{s1}, \text{s2} \}$:
Quantifying this phenomenon is straightforward: we stipulate that the hub neuron's spiking frequency -- denoted $\omega_{\text{hub}}(\mathbf{x})$ is close to an intermediate frequency of 0.55Hz.  
Mathematically, we achieve this via constraints on the mean and variance of the hub neuron spiking frequency.

@@

Third, we perform emergent property inference: we find a distribution over parameter configurations $\mathbf{z}$, and insist that samples from this distribution produce the emergent property; in other words, they obey the constraints introduced in Equation \ref{eq:EP_STG}.  
This distribution will be chosen from a family of probability distributions $\mathcal{Q} = \left\{ q_{\bm{\theta}}(\mathbf{z}) : \bm{\theta} \in \Theta \right\}$, defined by a deep generative distribution of the normalizing flow class \cite{rezende2015variational, dinh2016density, papamakarios2017masked} -- neural networks which transform a simple distribution into a suitably complicated distribution (as is needed here).  
This deep distribution is represented in Figure \ref{fig:STG}C (see Section \ref{methods_EPI}).  
Then, mathematically, we must solve the following optimization program: 
%\begin{equation} \label{eq:EPI}
%\begin{split}
%&\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
%&\text{  s.t.  } \mathbb{E}_{\mathbf{z} \sim q_{\bm{\theta}}}\left[ \mathbb{E}_{\mathbf{x} \sim p(\mathbf{x} \mid \mathbf{z})}\left[T(\mathbf{x})\right] \right] = \bm{\mu}, \\
%\end{split}
%\end{equation}
\begin{equation} \label{eq:EPI}
\begin{split}
q_{\bm{\theta}}(\mathbf{z} \mid \mathcal{X}) ~~=~~ &\argmax_{q_{\bm{\theta}} \in \mathcal{Q}} H(q_{\bm{\theta}}(\mathbf{z})) \\
&\text{  s.t.  } ~\mathcal{X}:~~ \mathbb{E}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\mu}, \text{Var}_{\mathbf{z},\mathbf{x}}\left[f(\mathbf{x}; \mathbf{z})\right] = \bm{\sigma}^2 \\
\end{split}
\end{equation}
%where $T(\mathbf{x}), \bm{\mu}$ are defined as in Equation \ref{eq:EP}, and $p(\mathbf{x} | \mathbf{z})$ is the intractable distribution of data from the model, $\mathbf{x}$, given that model's parameters $\mathbf{z}$ (we access samples from this distribution by running the model forward).
where $f(\mathbf{x}, \mathbf{z}), \bm{\mu}$, and $\bm{\sigma}$ are defined as in Equation \ref{eq:EP}.
According to the emergent property of interest, $f(\mathbf{x}, \mathbf{z})$ may contain multiple statistics, in which case the mean and variance vectors $\bm{\mu}$ and $\bm{\sigma}^2$ match this dimension.
%The purpose of each element in this program is detailed in Figure \ref{fig:STG}D.
Finally, we recognize that many distributions in $\mathcal{Q}$ will respect the emergent property constraints, so we select that which has maximum entropy.
This principle, captured in Equation \ref{eq:EPI} by the primal objective $H$, identifies parameter distributions with minimal assumptions beyond some chosen structure \cite{jaynes1957information, elsayed2017structure, loaiza2017maximum, savin2017maximum}.
Such a normative principle of maximum entropy, which is also that of Bayesian inference, naturally fits with our scientific objective of reasoning about parametric sensitivity and robustness. 
The recovered distribution of EPI is as variable as possible along each parametric manifold such that it produces the emergent property.

@@

In relation to broader methodology, inspection of the EPI objective reveals a natural relationship to posterior inference.
%Specifically, EPI executes variational inference in an exponential family model, the sufficient statistics and mean parameter of which are defined by the emergent property statistics and values, respectively (see Section \ref{methods_VI}).
%Specifically, EPI executes variational inference in an exponential family model, the sufficient statistics and mean parameter of which are defined by $f(\mathbf{x})$, $\bm{\mu}$ and $\bm{\sigma}$ (see Section \ref{methods_VI}).
Specifically, EPI executes a novel variant of  Bayesian inference with a uniform prior and a gaussian likelihood on the emergent property statistic (see Section \ref{methods_exp_fam_lh}).
A key advantage of EPI over established Bayesian inference is that the predictions made my the inferred distribution are constrained to produce the specified emergent property.
%Equipped with this method, we may examine structure in such inferred distributions or make comparisons between inferred distributions at graded values of emergent property statistics.
Equipped with this method, we may examine structure in posterior distributions or make comparisons between posteriors conditioned at different levels of the same emergent property statistic.
In Sections \ref{results_V1} and \ref{results_SC}, we prove out the value of EPI by using it to investigate and produce novel insights into two prominent models in neuroscience. 
Subsequently in Section \ref{results_LRRNN}, we show EPI's superiority in parameter scalability and fidelity of the posterior predictive distribution by conditioning on stable amplification in low-rank RNNs.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{EPI reveals how noise across neural population types governs Fano factor in a stochastic inhibition stabilized network} \label{results_V1}

\begin{figure}
\vspace{-1cm}
\begin{center}
\includegraphics[scale=0.2]{figures/fig2/fig2.pdf}
\end{center}
\vspace{-1cm}
\caption{\small Emergent property inference of a stochastic stabilized supralinear network.
\textbf{A}.  Four-population model of primary visual cortex with excitatory (black), parvalbumin (blue), somatostatin (red), and VIP (green) neurons (excitatory and inhibitory projections filled and unfilled, respectively).   
Some neuron-types largely do not form synaptic projections to others ($|W_{\alpha_1, \alpha_2})| < 0.025$).
Each neural population receives a baseline input $\mathbf{h}_b$, and the E- and P-populations also receive a contrast-dependent input $\mathbf{h}_b$.
Additionally, each neural population receives a slow noisy input $\bm{\epsilon}$.
\textbf{B}. Responses of the deterministic smodel ($\bm{\epsilon}=\bm{0}$) to varying contrasts.  The response at 50\% contrast (dots) is the focus of our analysis.
\textbf{C}. Paradoxical response of the stochastic model to a small increase in input to the P-population.  
\textbf{D}. EPI posterior of noise parameteres $\mathbf{z}$ conditioned on realistic E-population Fano factors.
The posterior predictive distribution is show on the bottom-left.
and the mode of the distribution is starred.
\textbf{E}. (Top) Enlarged visualization of the $\sigma_E$-$\sigma_P$ marginal distribution of the posterior.
Each gray dot is a choice of $\sigma_P$, for which a constrained mode $z^*(\sigma_P, P)$ is chosen.
The arrows show the most sensitive dimensions of the Hessian evaluated at these modes.
(Bottom) Such sensitive dimensions of the Hessian (dots) are significantly more sensitive than randomly chosen dimensions (box and whiskers).
\textbf{F}. The Fano factor of the E-population is strongly correlated with each other neuron-type population.
\textbf{G}. Mean and standard deviation (across EPI posterior) of Fano factor of each neuron-type population at each level of contrast.
}
 \label{fig:V1}
\end{figure}

Dynamical models of excitatory (E) and inhibitory (I) populations with supralinear input-output function have succeeded in explaining a host of experimentally documented phenomena.
In a regime characterized by inhibitory stabilization of strong recurrent excitation, these models give rise to paradoxical responses \cite{tsodyks1997paradoxical}, selective amplification  \cite{goldman2009memory, murphy2009balanced}, surround suppression \cite{ozeki2009inhibitory} and normalization \cite{rubin2015stabilized}. 
Despite their strong predictive power, E-I circuit models rely on the assumption that inhibition can be studied as an indivisible unit. 
However, experimental evidence shows that inhibition is composed of distinct elements -- parvalbumin (P), somatostatin (S), VIP (V) --
composing 80\% of GABAergic interneurons in V1 \cite{markram2004interneurons, rudy2011three, tremblay2016}, and that these inhibitory cell types follow specific connectivity patterns (Fig. \ref{fig:V1}A) \cite{pfeffer2013inhibition}.
Recent theoretical advances \cite{litwin2016inhibitory, GarciaDelMolino2017, Chen2019},  have only started to address the consequences of this multiplicity in the dynamics of V1, strongly relying on linear theoretical tools. 
Here, we use EPI to characterize the properties of slow noise in a stochastic version of this model, which result in biologically realistic responses.

We considered the contrast response of a nonlinear dynamical V1 circuit model (Fig. \ref{fig:V1}A) with a state comprised of each neuron-type population's rate $\mathbf{x} = \left[x_E, x_P , x_S, x_V \right]^\top$.
Each population receives recurrent input $W \mathbf{x}$ from synaptic projections of effective connectivity $W$ and an external input $\mathbf{h}$, which determine the population rate via nonlinearity $\phi = \left[\right]^2_+$ (see Section \ref{methods_V1}).
The circuit model evolves form an initial condition $\mathbf{x}(0) \sim \mathcal{U}(\left[10, 25\right])$ with time constant $\tau = 1$ms according to a contrast-dependent input $\mathbf{h}$ and  slow noise $\bm{\epsilon}$ of time constant $\tau_{\text{noise}} = 5$ms.
This model is the stochastic stabilized supralinear network (SSSN) \cite{hennequin2018dynamical}  generalized to have inhibitory multiplicity 
\begin{equation}
    \tau \frac{d\mathbf{x}}{dt} = -\mathbf{x} +\phi(W\mathbf{x} + \mathbf{h} + \bm{\epsilon}).
\end{equation}
As contrast increases, input to the E- and P-populations increases  relative to a baseline input $\mathbf{h}_b$ via $\mathbf{h}_c$
\begin{equation}
\mathbf{h} = \mathbf{h}_b + c\mathbf{h}_c,
\end{equation}
where $h_{c,E}, h_{c,P}  > 0$ and $h_{c,S}, h_{c,V}  = 0$.
In this analysis, we fixed $W$,$\mathbf{h}_b$, and $\mathbf{h}_c$ to values fit to mean contrast responses in mice with the deterministic model \cite{palmigiano2020structure} ($\bm{\epsilon}=\mathbf{0}$, Fig. \ref{fig:V1}B, see Section \ref{methods_V1}).
At all contrasts, the E-population of this SSSN is unstable without recurrent inhibitory feedback.
At 50\% contrast, only the P-population exhibits the paradoxical effect (\ref{fig:V1}C, Fig. \ref{fig:V1_1}), so the network is P-stabilized.

The slow noise of the SSSN is an an Ornstein-Uhlenbeck process
\begin{equation}
\tau_{\text{noise}} d\epsilon_\alpha = -\epsilon_\alpha dt + \sqrt{2\tau_{\text{noise}}}\sigma_\alpha dB,
\end{equation}
parameterized by $\sigma_\alpha$, which can be different for each neuron type,
\begin{equation}
\mathbf{z} = [\sigma_E, \sigma_P, \sigma_S, \sigma_V]^\top.
\end{equation} 
For this SSSN, we are interested in the parameters of slow noise that produce realistic stochastic fluctuations.
Here, we quantify this emergent property as having an excitatory population Fano factor near 1:
\begin{equation}\label{eq:EP_V1}
\begin{split}
\mathcal{X} ~~:~~  \mathbb{E}_{\mathbf{z}}\begin{bmatrix} F_E(\mathbf{x}; \mathbf{z}) \end{bmatrix}  &~~=~~ 1 \\ 
 \text{Var}_{\mathbf{z}}\begin{bmatrix} F_E(\mathbf{x}; \mathbf{z}) \end{bmatrix}  &~~=~~  0.125^2,
\end{split}
\end{equation}
where $F_\alpha(\mathbf{x}; \mathbf{z})$ is the Fano factor of the $\alpha$-population.
